Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-07 paper/paper.tex
Updated paper with subsection on handling 2bar

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage {algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

We now calculate a 0-CFA-style summary liveness for each GC point of a
program.  There are two parts  to this.  First, for each function $f$,
we determine a {\em  summary demand} $\sigma_f$ over-approximating any
demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
\cup \deltacall{f}{\pi^k}{g^k}$\comment{Rewrite, copied from earlier paper}.

@@

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The strongly regular CFGs obtained after the Mohri-Nederhof
transformation are
  first converted  into  NFAs.  The        algorithm         described        in
Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow}
\lang{\nt}$, where {\nt} accepts forward paths only.
\comment{Rewrite the algorithm to handle \cbar and update the example automata}
Thus  \nt\  can  be  used  to  check  
membership of forward paths.

@@

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

To avoid analysing the body of a function we assumed a symbolic demand $\sigma$
and created a summary liveness for the function arguments and the variables 
in the function. To get the actual liveness we have to substitute the actual 
demands that are placed on the function. We calculate an over-approximation 
of the actual demand, a {\em summary demand} and calculate the liveness at each GC point inside the 
function. Even though we lose a bit of precision due to this, our 
experiments show that this is acceptable \comment{Is this claim true?}.
The {\em summary demand} is calculated by taking a union of the demand at 
every call site of a function. A function call in our language can only be of the
form $\pi\!\!:\!(\LET~x  \leftarrow (f\,y_1\,\ldots\,y_n)\ldots$. The demand on 
$f$ due to this function call will be the liveness of $x$ at the program 
point $\pi$. Denoting this demand as $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
site $\pi$  in function $g$) we can write the {\em summary demand} as 
$\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
\cup \deltacall{f}{\pi^k}{g^k}$.

\comment{
We now calculate a 0-CFA-style summary liveness for each GC point of a
program.  There are two parts  to this.  First, for each function $f$,
we determine a {\em  summary demand} $\sigma_f$ over-approximating any
demand $\sigma$ passed to $f$.  So, consider a function $g$ containing
a  call   site  $\pi$   to  $f$,  say   $\pi\!\!:\!(\LET~x  \leftarrow
(f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.   Let   the  demand  on   $g$  be
$\sigma_g$ and the  liveness of $x$ at $\pi_i$  be $\Lanv{i}{x}$. Then
the $\LET$ case of Figure~\ref{fig:live-judge} tells us this call site
contributes  $\Lanv{i}{x}$ to  the  demand $\sigma_f$  placed on  $f$.
Write $\deltacall{f}{\pi_i}{g}$  for this demand  (contributed by call
site $\pi$  in $g$).  Thus, supposing  the $k$ call  sites to function
$f$  are  $\pi^1$ (in  function  $g^1$)  \ldots  $\pi^k$ (in  function
$g^k$),  then  the  over-approximation  requirement on  $\sigma_f$  is
achieved by  taking $\sigma_f =  \deltacall{f}{\pi^1}{g^1} \cup \cdots
\cup \deltacall{f}{\pi^k}{g^k}$Rewrite, copied from earlier paper}.

@@

\subsubsection{Removing the $\cbar$ edges}
Once the NFA's are created, we need to remove the transitions on 
$\cbar$. This edge is replaced by an $\epsilon$ transition and the 
$\cbar$ edge is deleted. If the original grammar had a non-null 
$\sigma$ then the NFA reaches a final state throught this path 
otherwise it does not reach a final state.\comment{Need to rewrite this}
%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The strongly regular CFGs obtained after the Mohri-Nederhof
transformation are
  first converted  into  NFAs.  The        algorithm         described        in
Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow}
\lang{\nt}$, where {\nt} accepts forward paths only.
\comment{Rewrite the algorithm to handle \cbar and update the example automata}
Thus  \nt\  can  be  used  to  check  
membership of forward paths.

@@

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,

%% \begin{figure}
%% \begin{algorithm}[H]
%%  \KwData{this text}
%%  \KwResult{how to write algorithm with \LaTeX2e }
%%  initialization\;
%%  \While{not at end of this document}{
%%   read current\;
%%   \eIf{understand}{
%%    go to next section\;
%%    current section becomes this one\;
%%    }{
%%    go back to the beginning of current section\;
%%   }
%%  }
%%  \caption{How to write algorithms}
%% \end{algorithm}
%% \label{algo:LGC-Algo} \figrule
%% \end{figure}
We   provide   an   intuition    about   the   correctness   of   this
algorithm Figure~\ref{algo:LGC-Algo}. Consider the following facts about the print stack during a
GC,
